---
editor_options: 
  markdown: 
    wrap: 72
---

# Analysis of longitudinal data

*Describe the work you have done this week and summarize your learning.*

-   Describe your work and results clearly.
-   Assume the reader has an introductory course level understanding of
    writing and reading R code as well as statistical methods.
-   Assume the reader has no previous knowledge of your data or the more
    advanced methods you are using.

```{r}
date()

# load the GGally and ggplot2 libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(lme4)
```

## Part I

### Description of the RAT data

First of all, let's have a look at the data and give a brief description
of the dataset.

```{r}
# Read the RATS data
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt",
                   sep = "\t", header = T)

# Look at the (column) names of RATS
names(RATS)

# Look at the structure of RATS
str(RATS)

# Print out summaries of the variables
summary(RATS)
```

The RATS dataset in the wide data form: we have 16 subjects, rats, that
are split into three groups and biological measurements are taken for
each rat every week. Time points are **WD1**, **WD8**, **WD15**,
**WD22**, **WD29**, **WD36**, **WD43**, **WD44**, **WD50**, **WD57** and
**WD64**.

### Graphical Displays of Longitudinal Data

```{r}
# convert the ID and Group columns to factors
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)

# convert to longitudinal form
RATS_L <- pivot_longer(RATS, cols = -c(ID, Group), names_to = "WD",
                     values_to = "Weight") %>%
  mutate(Time = as.integer(substr(WD, 3,4))) %>%
  arrange(Time)

dim(RATS_L)
glimpse(RATS_L)
```

Now the dataset is in the longitudinal form. In the wide form, we had
one line per each rat where variables were the groups and the
time-points when the biological measurements were taken. However, in the
longitudinal form, we have as many lines as many rats \* time points
(and 16 \* 11): each line now is rat #1 - time point #1 - biological
measurement for time point #1.

```{r}
# draw the plot
ggplot(RATS_L, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATS_L$Weight), max(RATS_L$Weight)))
```

From this first plot, we can already appreciate that the weight of
almost all the rats tend to increase during the treatment. Also, we can
observe the tracking phenomenon (rats whose weight was high at the
beginning weight more throughout the study).\
However, the non-standardized nature of the data makes drawing
conclusions hard.

```{r}
# standardise the variable Weight
RATS_L_st <- RATS_L %>%
  group_by(Time) %>%
  mutate(Weight_st = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()
# look at the data
glimpse(RATS_L_st)
# plot again with the standardised Weight
ggplot(RATS_L_st, aes(x = Time, y = Weight_st, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized weight")


# summary data with mean and standard error of weight by treatment and week 
RATS_L_st_se <- RATS_L_st %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = (sd(Weight)/sqrt(Weight)) ) %>%
  ungroup()

glimpse(RATS_L_st_se)

# plot the mean profiles
ggplot(RATS_L_st_se, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.5)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

From this last plot, showing average profiles for each rats group with
mean and standard error at each time point, we can see that there is no
overlap in the means of the rat groups.

### Another visualization

```{r}
# create a summary data by Group and ID with mean as the summary variable
# (ignoring baseline Time 1)
RATS_S <- RATS_L %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

glimpse(RATS_S)

# draw a boxplot of the mean versus Group
ggplot(RATS_S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Days 1-64")
```

The created boxplots reflect quite well what we observed from the
previous plots. This plot also shows that the variability of Group 3 is
the smallest. All the groups are characterized by the presence of an
outlier each: we can try to remove them.

```{r}
# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
RATS_S_clean <- RATS_S[!(RATS_S$Group == 3 & RATS_S$mean < 500),] %>%
  subset(mean < 590) %>% subset(mean > 240)

# plot again
ggplot(RATS_S_clean, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Days 1-64")
```

### ANOVA

Since t-test require 2 levels (Groups) and we have 3 of them, such test
can't be applied. Let's move to ANOVA instead

```{r}
# Add the RATS_S from the original data as a new variable to the summary data
RATS_S <- RATS_S %>%
  mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response 
fit1 <- lm(mean ~ baseline + Group, data = RATS_S)
summary(fit1)
# Compute the analysis of variance table for the fitted model with anova()
anova(fit1)
```
Looking at the model and at the ANOVA testing, when the baseline is taken into
account, the groups don't seem to be significantly different anymore. In other
words, the baseline is  explaining most of variability between the two groups.

## Part II

### Description of the BPRS data

First of all, let's have a look at the data and give a brief description
of the dataset.

```{r}
# Read the BPRS data
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt",
                   sep = " ", header = T)

# Look at the (column) names of RATS
names(BPRS)

# Look at the structure of RATS
str(BPRS)

# Print out summaries of the variables
summary(BPRS)
```

The BPRS dataset in the wide data form: we have 40 subjects, men, that
are split into two groups and biological measurements are taken for each
man every week, from week0 (baseline measurement) to week8.

### Convert to longitudinal form

```{r}
# convert categorical variables to factors
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)
# convert to longitudinal form
BPRS_L <-  pivot_longer(BPRS, cols= -c(treatment,subject), names_to = "weeks",
                       values_to = "bprs") %>% arrange(weeks)
BPRS_L <-  BPRS_L %>% mutate(week = as.integer(substr(weeks,5,5)))
```

Now the dataset is in the longitudinal form. In the wide form, we had
one line per each man where variables were the groups and the
time-points when the biological measurements were taken. However, in the
longitudinal form, we have as many lines as many men \* time points (and
40 \* 9): each line now is man #1 - time point #1 - biological
measurement for time point #1.

### Visualise the data

```{r}
# plot the RATSL data
ggplot(BPRS_L, aes(x = week, y = bprs, group = interaction(subject, treatment))) +
  geom_line(aes(linetype = treatment )) + 
  scale_y_continuous(name = "bprs") + 
  theme(legend.position = "top")
```

Looking at the plot, it looks like there's no difference in the effect
of the two different treatments since the lines from the two groups
overlap quite well.

### Linear model

```{r}
# create a regression model RATS_reg
fit_BPRS <- lm(bprs ~ week + treatment, data = BPRS_L)

# print out a summary of the model
summary(fit_BPRS)
```

From the fitted linear model, we can see that time (week variable) is
very significant in the regression. Treatment2 doesn't look important
though. However, the independence assumption is violated, so we would need a random
intercept model.

### The Random Intercept Model
As explained before, the linear model assumes independence of the repeated measures
of bprs, but it's not the case: we need to fit a random intercept model for the
same two explanatory variables, week and treatment. This will allow the
linear regression fit for each man to differ in intercept from other men.

```{r}
library(lmerTest)
# Create a random intercept model
fit_BPRS_2 <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRS_L,
                   REML = FALSE)

# Print the summary of the model
summary(fit_BPRS_2)
```

From the Random Intercept Model we can have a look at the random effects. The
**subject** derived effect looks to be very important in the model. This can be
observed by the fact that t.test is given by estimate (variance, in this case
104.21) / SD (SD = SE/sqrt(sample size)): if the resulting t.test is > abs(1.93),
then the pvalue is significant and the effect is significant as well. \
by running this model wich takes into account the randomness coming from the 
**subject**, we correct the fixed effects and now we can trust these results

### Slippery slopes: Random Intercept and Random Slope Model

Now we can move on to fit the random intercept and random slope model.
Fitting a random intercept and random slope model allows the linear
regression fits for each individual to differ in intercept but also in
slope.

```{r}
fit_BPRS_3 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRS_L,
                   REML = FALSE)

# print a summary of the model
summary(fit_BPRS_3)

# perform an ANOVA test on the two models
anova(fit_BPRS_3, fit_BPRS_2)
```
When adding **week** as a random effect, we improve the model (anova testing results
in a pvalue < 0.05). This means that **week** is important as well as random
effect.


### Random Intercept and Random Slope Model with interaction

```{r}
# create a random intercept and random slope model with the interaction
fit_BPRS_interaction <- lmer(bprs ~ week * treatment + (week | subject), data = BPRS_L,
                   REML = FALSE)

# print a summary of the model
summary(fit_BPRS_interaction)

# perform an ANOVA test on the two models
anova(fit_BPRS_interaction, fit_BPRS_3)

# plot the RATSL data
ggplot(BPRS_L, aes(x = week, y = bprs, group = interaction(subject, treatment))) +
  geom_line(aes(linetype = treatment )) + 
  scale_y_continuous(name = "bprs") + 
  theme(legend.position = "top")

# Create a vector of the fitted values and add to BPRS_L
Fitted <- fitted(fit_BPRS_interaction)
BPRS_L$fitted <- Fitted

# plot the RATSL data
ggplot(BPRS_L, aes(x = week, y = fitted, group = interaction(subject, treatment))) +
  geom_line(aes(linetype = treatment )) + 
  scale_y_continuous(name = "bprs") + 
  theme(legend.position = "top")
```
Finally, when we fit a model with the interaction between **week** and **treatment**
we get no significant difference from the previous one (anova pvalue > 0.05).
